---
title: "data_processing_notes"
output: html_document
date: '2022-07-13'
---
 ## Introduction to dplyr 
 dplyr is an R pakage in the tidyverse. We can load the package using:
```{r}
install.packages("dplyr")
library(dplyr)
?dplyr
library(ggplot2)
head(starwars)
```
### Filtering Data 
We can filter data using  `filter()` which allows for us to subset based on row value
Tips:

* Be sure to speell column name right and the valye name if needed. R is case sensitive 
* Be sure to use `==` when comparing observations. 
* Use >, <, <=, >= to compare numeric or categorical variables (nominal variables are ranked alphabetically, while ordinal variables have a built-in rank)
```{r}
filter(starwars, hair_color == "blond" | hair_color == "blonde")
filteredData <- filter(starwars, skin_color == "fair", species == "Human", homeworld == "Tatooine")
places <- c("Tatooine", "Naboo")
filter(starwars, homeworld %in% places)
```
Diamonds 
```{r}
head(diamonds) # displays first 6 rows
class(diamonds$cut) # gives specific type of data 
summary(diamonds$cut) # count or summary statistic 
summary(diamonds$carat)
head(diamonds$cut) # displays first 6 

filter(diamonds, cut > "Good")

#ordering categorical data
unique(starwars$eye_color)
factor(starwars, c("red", "orange", "gold", "yellow", "green, yellow", "blue"), ordered = T)

# find all chracters that are shorter than 100 cm 
head(starwars)
filter(starwars, height < 100, species != "Droid")

#find all characters who were born between the years 100 and 200 
filter(starwars, birth_year >= 100 & birth_year <= 200)

# find all character with weight over 100 but are shorter than 185 
filter(starwars, mass > 100 & height < 185)

#Missing values 

NA > 5 # doesn't know 

# find which characters are missing a hair color 
filter(starwars, is.na(hair_color))
```

### Arranging Data 
`arrange()` reorders rows. It does not remove any rows. NA values are always at the end when you order by a column
```{r}
arrange(starwars, desc(birth_year))

arrange(starwars, hair_color, birth_year)

# Star wars tallest and shortest 
arrange(starwars, desc(height))
# A-Z by name
arrange(starwars, name)
# How to use arrange() to sort all missing values to the start 
arrange(starwars, desc(is.na(starwars)))
```

### We can selct certian columns in the dataset 
`select()` allows us to retain only certain variables (columns).It doesn't change the order but it removes coloumns not named
```{r}
select(starwars, name, birth_year)
select(starwars, contains("_"))

# Practice 1 
# Select out the homeworld and species for the data set 
select(starwars, homeworld, species)
# How to select hair_color, eye_color, skin_color and birth_year
select(starwars, ends_with("color"), birth_year)
select(starwars, contains("_"))
select(starwars, starts_with("h"), starts_with("e"), starts_with("s"), starts_with("b"))
```

### Adding new columns 
`mutate()` adds new columns to the end of your dataset
```{r}
small <- select(starwars, height, mass, birth_year)
head(small)
small <- mutate(small, height_m = height/100)
mutate(small, bmi = mass/(height_m^2))
```
###Piping 
```{r}
species_masses <- summarise(group_by(starwars, species),
mean_mass = mean(mass, na.rm = 1),
count = nO) # nO returns the number of values in each species
# the pipe operator %>%
# function(x, y) is the same as x %>% function(y)
# When using dplyr functions, generally always start with dataset
species_masses <- starwars %>% 
  group_by(species) %>%
  summarise(mean_mass = mean(mass, na.rm = T), count = nO)
arrange(species_masses, desc(mean_mass))
```

### Sampling a designated number of rows
`sample_n()` allows us to sample a random number of rows from our dataset
```{r}
starwars_10rows <- sample_n(starwars,10)
starwars_10rows
# 10% of rows
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent
```

## Slicing 
We can take a "slice" of a dataset using `slice()` and its related set of functions. Retains rows and is not random.
```{r}
slice(starwars, 50:50) # slice by index value (row number)
slice_head(starwars, n = 10) # first 10 rows by index 
slice_tail(starwars, n = 10) # last 10 rows by  index 
slice_min(starwars, order_by = height, n = 10) # rows with losed 10 heigths 
slice_max(starwars, order_by = height, n = 10) # rows with 10 tallest

```